### 2020-05-29 writeup/transfer.tex
fig 1
sec 4

\textbf{Model dissimilarity.}
We measure model dissimilarity as $\norm{\beta_1 - \beta_2}^2$, which is the distance between source and target in the isotropic model.
We derive a sharp threshold when positive transfer transitions to negative transfer, as model dissimilarity increases.
%Based on Theorem \ref{thm_model_shift}, we derive the transition threshold in the following proposition.
Define the following function
\begin{align*}
	\Phi(\rho_1, \rho_2) = \frac{(\rho_1 + \rho_2 - 1)^2}{\rho_1 (\rho_1 + \rho_2) (\rho_2 - 1)}.
\end{align*}

\begin{proposition}\label{prop_dist_transition}
	In the isotropic model, assume that $\rho_1 > 40$ is a fixed constant.
	Let $\sigma_1 = \sigma_2 = \sigma$.
	Whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ is determined by the ratio between model dissimilarity and $\Phi(\rho_1, \rho_2)$:
	\begin{itemize}
		\item If ${p d^2} < \frac{1}{2}  {\sigma^2}  \cdot \Phi(\rho_1, \rho_2)$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If ${p d^2} \ge 2 {\sigma^2} \cdot \Phi(\rho_1, \rho_2)$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})$.
	\end{itemize}
\end{proposition}

@@

\begin{figure}
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/model_shift_phase_transition.eps}
		\caption{Model dissimilarity}
		\label{fig_model_shift}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/datapoints_phase_transition.eps}
		\caption{Large source task data size}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/complementary.eps}
		\caption{Covariate shift}
		\label{fig_covariate}
	\end{subfigure}
	\caption{Three contributing causes of negative transfer:
	(a) As model dissimilarity increases, we observe a transition from positive to negative transfer  (Proposition \ref{prop_dist_transition}).
	(b) As source task data size increases, we observe a transition positive to negative transfer (Proposition \ref{prop_data_size}).
	(c) As covariate shift becomes more severe, test performance gets worse (Proposition \ref{prop_covariate}).}
	\label{fig_model_shift_phasetrans}
\end{figure}

\textbf{Data size.}
In classical Rademacher or VC based theory of multi-task learning, the generalization bounds are usually presented for settings where the data sizes are equal for all tasks \cite{B00,M06,MPR16}.
More generally, such results are still applicable when all task data are being added simultaneously.
On the other hand, for many applications of multi-task learning, the data sources are usually heterogeneous.
For such settings, we have observed that adding more labeled data from one task does not always help.
%On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.
Using the isotropic model, we consider what happens if we vary the source task data size.

\begin{proposition}\label{prop_data_size}
	In the isotropic model, assume that $\rho_1 > 40$ and $\rho_2 > 500$ are fixed constants.
	We have the following conditions to determine whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$:
	\begin{itemize}
		\item[a)] If $pd^2 \le \frac{1}{2}\cdot \frac{\sigma^2}{\rho_2-1}$, then whp $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$ for any $\rho_1 > 40$.
		\item[b)] If $pd^2 > 2\cdot\frac{\sigma^2}{\rho_2 - 1}$, then we have the following transition depending on $\rho_1$:
		\begin{itemize}
			\item If $\rho_1 \cdot p < \frac{(\rho_2-2)\sigma^2}{(1 + {\rho_1}^{-0.5})^4(\rho_2 - 1) d^2 - \sigma^2 / p}$, then whp $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
			\item If $\rho_1 \cdot p > \frac{(\rho_2-2) \sigma^2}{(1 - {\rho_1}^{-0.5})^4 (\rho_2 - 3) d^2 - \sigma^2 / p}$, then whp $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
		\end{itemize}
	\end{itemize}
\end{proposition}
The intuition behind Proposition \ref{prop_data_size} is as follows.
a) If $\norm{\beta_1 - \beta_2}^2 \approx pd^2$ is sufficiently small, then adding any amount of labeled data from the source task always provides positive transfer;
ii) Otherwise, as source data size increases, we observe a transition from positive to negative transfer.
The proof of Proposition \ref{prop_data_size} is similar to Proposition \ref{prop_dist_transition}, and can be found in Appendix \ref{app_proof_31}.

\textbf{Covariate shift.}
So far we have considered the isotropic model where $\Sigma_1 = \Sigma_2$.
This setting is relevant for settings where different tasks share the same input features such as multi-class image classification.
In general, the covariance matrices of the two tasks may be different such as in text classification.
In this part, we consider what happens when $\Sigma_1 \neq \Sigma_2$.

@@

\begin{proposition}\label{prop_covariate}
	In the setting described above, assume that $\rho_1 > 3$ and $\rho_2>3$,
and $\|\Sigma_1\|\le C_1$ for some constant $C_1>0$.
	Let $g(M)$ denote the test error of $\hat{\beta}_t^{\MTL}$ when the covariance shift matrix is equal to $M\in\cS_{\mu}$.
	We have that \[ g(\mu\id) \le \bigbrace{1+ \bigo{\frac{\rho_2}{\rho_1} + \frac{1}{\sqrt {\rho_1}}}} \min_{M\in\cS_{\mu}} g(M). \]
\end{proposition}
Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

@@

\textbf{Labeled data efficiency.}
We use our tools to explain a key result of taskonomy \cite{ZSSGM18}, which shows that by learning from multiple related tasks, one can reduce the amount of labeled data from each task.
This is formalized by a metric called the data efficiency ratio as follows.

@@

\begin{proposition}\label{prop_data_efficiency}
	In the isotropic model, assume that $\rho_1$ and $\rho_2$, $pd^2 < {8\sigma^2} /{(3 \rho)}$.
	Then the data efficiency ratio is at most $\frac{1}{2\rho} + \frac{\sigma^2}{2\sigma^2 - 3p d^2 \rho / 4}$.
\end{proposition}
The proof of Proposition \ref{prop_data_efficiency} can be found in Appendix \ref{app_proof_data}.

\textbf{Labeled data de-noising.}
We further show that multi-task learning is particular powerful when the labeled data of the source task is less noisy compared to the target task.
Consider a more general setting of Proposition \ref{prop_dist_transition} where the noise level $\sigma_1$ of task $1$ differs from the noise level $\sigma_2$ of task $2$.
We extend Proposition \ref{prop_dist_transition} as follows.

\begin{proposition}\label{prop_var_transition}
	In the isotropic model, assume that $\rho_1 > 40$ and $pd^2 < \frac{1}{2} {\sigma_2^2}  \cdot \Phi(\rho_1, \rho_2)$.
	We derive the following transition as a parameter of $\sigma_1^2$:
	\begin{itemize}
		\item If $\sigma_1^2 \le -2 \rho_1 \cdot p d^2+\left(1+ \frac34\rho_1 \Phi(\rho_1, \rho_2)\right)\cdot\sigma_2^2$, then whp $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\sigma_1^2 > - \frac12\rho_1\cdot p d^2   +\left(1+ \frac32\rho_1\Phi(\rho_1, \rho_2)\right) \cdot \sigma_2^2$, then whp $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\end{itemize}
\end{proposition}
As a corollary, if $\sigma_1^2 \le \sigma_2^2$, then we always get positive transfer.
The proof of Proposition \ref{prop_var_transition} is similar to Proposition \ref{prop_dist_transition}.
The details can be found in Appendix \ref{app_proof_data}.

>>>>

\textbf{Cause 1.}
We measure model dissimilarity as $\norm{\beta_1 - \beta_2}^2$, which is the distance between source and target in the isotropic model.
We derive a sharp threshold when positive transfer transitions to negative transfer, as model dissimilarity increases.
%Based on Theorem \ref{thm_model_shift}, we derive the transition threshold in the following proposition.
Define the following function
\begin{align*}
	\Phi(\rho_1, \rho_2) = \frac{(\rho_1 + \rho_2 - 1)^2}{\rho_1 (\rho_1 + \rho_2) (\rho_2 - 1)}.
\end{align*}

\begin{proposition}[Model dissimilarity]\label{prop_dist_transition}
	In the isotropic model, assume that $\rho_1 > 40$ is a fixed constant.
	Let $\sigma_1 = \sigma_2 = \sigma$.
	Whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ is determined by the ratio between model dissimilarity and $\Phi(\rho_1, \rho_2)$:
	\begin{itemize}
		\item If ${p d^2} < \frac{1}{2}  {\sigma^2}  \cdot \Phi(\rho_1, \rho_2)$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If ${p d^2} \ge 2 {\sigma^2} \cdot \Phi(\rho_1, \rho_2)$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})$.
	\end{itemize}
\end{proposition}

@@

\begin{figure}
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/model_shift_phase_transition.eps}
		\caption{Model dissimilarity}
		\label{fig_model_shift}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/datapoints_phase_transition.eps}
		\caption{Large source task data size}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/complementary.eps}
		\caption{Covariate shift}
		\label{fig_covariate}
	\end{subfigure}
	\caption{Three contributing causes of negative transfer:
	(a) As model dissimilarity increases, we observe a transition from positive to negative transfer  (Proposition \ref{prop_dist_transition}).
	(b) As source task data size increases, we observe a transition positive to negative transfer (Proposition \ref{prop_data_size}).
	(c) As covariate shift becomes more severe (measured by increasing $\lambda$), test performance gets worse (Proposition \ref{prop_covariate}).}
	\label{fig_model_shift_phasetrans}
\end{figure}

\textbf{Cause 2.}
In classical Rademacher or VC based theory of multi-task learning, the generalization bounds are usually presented for settings where the data sizes are equal for all tasks \cite{B00,M06,MPR16}.
More generally, such results are still applicable when all task data are being added simultaneously.
On the other hand, for many applications of multi-task learning, the data sources are usually heterogeneous.
For such settings, we have observed that adding more labeled data from one task does not always help.
%On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.
Using the isotropic model, we consider what happens if we vary the source task data size.

\begin{proposition}[Source data size]\label{prop_data_size}
	In the isotropic model, assume that $\rho_1 > 40$ and $\rho_2 > 500$ are fixed constants.
	We have the following conditions to determine whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$:
	\begin{itemize}
		\item[a)] If $pd^2 \le \frac{1}{2}\cdot \frac{\sigma^2}{\rho_2-1}$, then whp $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$ for any $\rho_1 > 40$.
		\item[b)] If $pd^2 > 2\cdot\frac{\sigma^2}{\rho_2 - 1}$, then we have the following transition depending on $\rho_1$:
		\begin{itemize}
			\item If $\rho_1 \cdot p < \frac{(\rho_2-2)\sigma^2}{(1 + {\rho_1}^{-0.5})^4(\rho_2 - 1) d^2 - \sigma^2 / p}$, then whp $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
			\item If $\rho_1 \cdot p > \frac{(\rho_2-2) \sigma^2}{(1 - {\rho_1}^{-0.5})^4 (\rho_2 - 3) d^2 - \sigma^2 / p}$, then whp $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
		\end{itemize}
	\end{itemize}
\end{proposition}
The intuition behind Proposition \ref{prop_data_size} is as follows.
a) If $\norm{\beta_1 - \beta_2}^2 \approx pd^2$ is sufficiently small, then adding any amount of labeled data from the source task always provides positive transfer;
ii) Otherwise, as source data size increases, we observe a transition from positive to negative transfer.
The proof of Proposition \ref{prop_data_size} is similar to Proposition \ref{prop_dist_transition}, and can be found in Appendix \ref{app_proof_31}.

\textbf{Cause 3.}
So far we have considered the isotropic model where $\Sigma_1 = \Sigma_2$.
This setting is relevant for settings where different tasks share the same input features such as multi-class image classification.
In general, the covariance matrices of the two tasks may be different such as in text classification.
In this part, we consider what happens when $\Sigma_1 \neq \Sigma_2$.

@@

\begin{proposition}[Covariate shift]\label{prop_covariate}
	In the setting described above, assume that $\rho_1 > 3$ and $\rho_2>3$,
and $\|\Sigma_1\|\le C_1$ for some constant $C_1>0$.
	Let $g(M)$ denote the test error of $\hat{\beta}_t^{\MTL}$ when the covariance shift matrix is equal to $M\in\cS_{\mu}$.
	We have that \[ g(\mu\id) \le \bigbrace{1+ \bigo{\frac{\rho_2}{\rho_1} + \frac{1}{\sqrt {\rho_1}}}} \min_{M\in\cS_{\mu}} g(M). \]
\end{proposition}
Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

@@

\textbf{Implication 1.}
We use our tools to explain a key result of taskonomy \cite{ZSSGM18}, which shows that by learning from multiple related tasks, one can reduce the amount of labeled data from each task.
This is formalized by a metric called the data efficiency ratio as follows.

@@

\begin{proposition}[Labeled data efficiency]\label{prop_data_efficiency}
	In the isotropic model, assume that $\rho_1$ and $\rho_2$, $pd^2 < {8\sigma^2} /{(3 \rho)}$.
	Then the data efficiency ratio is at most $\frac{1}{2\rho} + \frac{\sigma^2}{2\sigma^2 - 3p d^2 \rho / 4}$.
\end{proposition}
The proof of Proposition \ref{prop_data_efficiency} can be found in Appendix \ref{app_proof_data}.

\textbf{Implication 2.}
We further show that multi-task learning is particular powerful when the labeled data of the source task is less noisy compared to the target task.
Consider a more general setting of Proposition \ref{prop_dist_transition} where the noise level $\sigma_1$ of task $1$ differs from the noise level $\sigma_2$ of task $2$.
We extend Proposition \ref{prop_dist_transition} as follows.

\begin{proposition}[Labeled data de-noising]\label{prop_var_transition}
	In the isotropic model, assume that $\rho_1 > 40$ and $pd^2 < \frac{1}{2} {\sigma_2^2}  \cdot \Phi(\rho_1, \rho_2)$.
	We derive the following transition as a parameter of $\sigma_1^2$:
	\begin{itemize}
		\item If $\sigma_1^2 \le -2 \rho_1 \cdot p d^2+\left(1+ \frac34\rho_1 \Phi(\rho_1, \rho_2)\right)\cdot\sigma_2^2$, then whp $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\sigma_1^2 > - \frac12\rho_1\cdot p d^2   +\left(1+ \frac32\rho_1\Phi(\rho_1, \rho_2)\right) \cdot \sigma_2^2$, then whp $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\end{itemize}
\end{proposition}
As a corollary, if $\sigma_1^2 \le \sigma_2^2$, then we always get positive transfer.
The proof of Proposition \ref{prop_var_transition} is similar to Proposition \ref{prop_dist_transition}.
The details can be found in Appendix \ref{app_proof_data}.
